Mergeable Types

### 2017-02-28 paper/implementation.tex
Typos.

Importantly, Irmin supports user-defined three-way merge functions for
reconciling concurrent operations. While Irmin's merge functions are
defined over objects on Irmin's content-addressable heap, \name's
merge functions are defined over OCaml types. We address this
representational mismatch with the help of OCaml's PPX metaprogramming
support~\cite{ppx} to derive bi-directional transformations between
objects on OCaml and Irmin heaps. We also derive the various
serialization functions required by Irmin\footnote{
  The Since \nameMonad monad relies on Irmin for realizing the branch
  structure, it is in fact parameterized over a mergeable type that
  implements all these functions. Thus the monad is functorized over
  the mergeable type, but not polymorphic. The convenience of type
  parameter inference that polymorphism offers can however be obtained
  for module parameters using OCaml extensioms, such as Modular
  Implicits~\cite{implicits}.
}.
Synchronization between replicas is performed using Git's notion of
pushing and pulling updates from remotes~\cite{git-tp}. As a result, we reap the
benefits of efficient delta-transfer (only missing objects are transferred between
replicas), compression and end-to-end encrypted communication between
the replicas.

>>>>

Importantly, Irmin supports user-defined three-way merge functions for
reconciling concurrent operations. While Irmin's merge functions are
defined over objects on Irmin's content-addressable heap, \name's
merge functions are defined over OCaml types. We address this
representational mismatch with the help of OCaml's PPX metaprogramming
support~\cite{ppx} to derive bi-directional transformations between
objects on OCaml and Irmin heaps. We also derive the various
serialization functions required by Irmin\footnote{
  The Since \nameMonad monad relies on Irmin for realizing the branch
  structure, it is in fact parameterized over a mergeable type that
  implements all these functions. Thus the monad is functorized over
  the mergeable type, but not polymorphic. The convenience of type
  parameter inference that polymorphism offers can however be obtained
  for module parameters using OCaml extensions, such as Modular
  Implicits~\cite{implicits}.
}.
Synchronization between replicas is performed using Git's notion of
pushing and pulling updates from remotes~\cite{git-tp}. As a result, we reap the
benefits of efficient delta-transfer (only missing objects are transferred between
replicas), compression and end-to-end encrypted communication between
the replicas.

### 2017-02-28 paper/meta.tex
Typos.

Fortunately, unlike source control systems where branching structure
is entirely dictated by the user, \name abstracts away branching
structure from the programmer, and hence retains the ability to
manifest it in a way that it deems fit. In particular, \name solves
the problem of multiple LCAs by suitably constraining the branching
structure such that the problem never arises. The constraints are
imposed either implicitly, as a result of how the operational semantics
defines an atomic step, or explicitly, by insisting that certain
conditions be met before merging a pair of versions
(\rulelabel{E-Pull-Wait}). First, the operational semantics already
disables criss-cross merges since it only ever merges versions that
are latest on their respective branches. Second, we impose certain
pre-conditions on the merging branches to preempt the structure shown
in Fig.~\ref{fig:external-lcas}. The intuition is as follows: consider
the branch $b_3$ at the instance of merging $v_{12}$. Since it has
already merged $v_{22}$, $v_{22}$ could be a common ancestor for $b_3$
and some other branch (call it $b$). Now, if $b_3$ merges $v_{12}$,
the same could be true of $v_{12}$. Since $v_{12}$ and $v_{22}$ are not
ordered by the ancestor relation, both become LCAs of
$b_3$ and $b$. We observe that this scenario can be prevented if, when
merging $b_1$, $b_3$ insists on an ancestor relation between the last
merged version ($v_{22}$) and the currently merging version.  We call
the last merged version of a branch its \emph{locus}. By maintaining a
total order of among locii of a branch $b$ as new versions are
created, we effectively guarantee the presence of a version (the
locus) that is lower than all the ancestors of the latest version on
$b$. Next, by requiring the locus to be an ancestor of the merging
version, we ensure that all the common ancestors have a single lowest
element, thus enforcing the uniqueness of the LCA.

@@

The proof is by the induction on the size of branching history. The
key observation is that the smallest history that is the union of
causal histories of two unmergeable versions (call them $v_i$ and
$v_j$) is smaller than the history that includes the unmergeable
versions. Since causal histories have maximum versions (respective
locii), which, by the inductive hypothesis, can be made mergeable, we
merge them to yield a new version $v$ that is the maximum of the
causal histories of both $v_i$ and $v_j$. If $v_i$ and $v_j$ are the
latest versions on their respective branches, the branches are now
mergeable.

>>>>

Fortunately, unlike source control systems where branching structure
is entirely dictated by the user, \name abstracts away branching
structure from the programmer, and hence retains the ability to
manifest it in a way that it deems fit. In particular, \name solves
the problem of multiple LCAs by suitably constraining the branching
structure such that the problem never arises. The constraints are
imposed either implicitly, as a result of how the operational semantics
defines an atomic step, or explicitly, by insisting that certain
conditions be met before merging a pair of versions
(\rulelabel{E-Pull-Wait}). First, the operational semantics already
disables criss-cross merges since it only ever merges versions that
are latest on their respective branches. Second, we impose certain
pre-conditions on the merging branches to preempt the structure shown
in Fig.~\ref{fig:external-lcas}. The intuition is as follows: consider
the branch $b_3$ at the instance of merging $v_{12}$. Since it has
already merged $v_{22}$, $v_{22}$ could be a common ancestor for $b_3$
and some other branch (call it $b$). Now, if $b_3$ merges $v_{12}$,
the same could be true of $v_{12}$. Since $v_{12}$ and $v_{22}$ are not
ordered by the ancestor relation, both become LCAs of
$b_3$ and $b$. We observe that this scenario can be prevented if, when
merging $b_1$, $b_3$ insists on an ancestor relation between the last
merged version ($v_{22}$) and the currently merging version.  We call
the last merged version of a branch its \emph{locus}. By maintaining a
total order of among loci of a branch $b$ as new versions are
created, we effectively guarantee the presence of a version (the
locus) that is lower than all the ancestors of the latest version on
$b$. Next, by requiring the locus to be an ancestor of the merging
version, we ensure that all the common ancestors have a single lowest
element, thus enforcing the uniqueness of the LCA.

@@

The proof is by the induction on the size of branching history. The
key observation is that the smallest history that is the union of
causal histories of two unmergeable versions (call them $v_i$ and
$v_j$) is smaller than the history that includes the unmergeable
versions. Since causal histories have maximum versions (respective
loci), which, by the inductive hypothesis, can be made mergeable, we
merge them to yield a new version $v$ that is the maximum of the
causal histories of both $v_i$ and $v_j$. If $v_i$ and $v_j$ are the
latest versions on their respective branches, the branches are now
mergeable.
